---
title: "Homework 1 Solution"
subtitle: ""
author: ""
date: ""
output: pdf_document
---


# Question 1

Here is a one possible solution. 

We first set up parameters and load relevant libraries.

```{r}
rm(list=ls())
library(data.table)
library(kknn)

# Primitives
N_train=100
N_test=10000
```

Next, we write functions for generating data and plotting.

```{r}
# Function to simulate data
Simulate <- function(func, N_train, N_test){
  set.seed(11)
  n=N_train+N_test
  x=runif(n, -1, 1)
  y=func(x)+rnorm(n, mean=0, sd=0.15)
  dt=data.table(y=y,x=x)
  tr_idx=sample(1:n, N_train)
  train=dt[tr_idx,]
  setkey(train,x)
  test=dt[-tr_idx,]
  setkey(test,x)
  return(list(model=func, full=dt, train=train, test=test))
}

# Function to create base scatter plot
Plot <- function(sim, method){
  # unpack
  true_model=sim$model
  train=sim$train
  if (method=="linear regression"){
    plot(train$x, train$y, cex=0.5, pch=19, col="grey")
    title(main="Scatter plot, true relationship, and linear fit")
    lines(train$x, true_model(train$x), col="black", lwd=2)
    linear=lm(y~x, data=train)
    yhat=predict(linear, train)
    lines(train$x, yhat, col="blue", lty=2, lwd=3)
    return(linear)
  } else if (method=="k-nn") {
    k_vec=2:15
    #near_list <- vector("list", length(k_vec))
    fit=matrix(0, nrow=dim(train)[1], ncol=length(k_vec))
    for (i in 1:length(k_vec)){
      k=k_vec[i]
      near <- kknn(y~x, train=train, test=train[,.(x)],k=k, kernel='rectangular')
      #near_list[[i]]=near
      fit[,i]=near$fitted
    }
    
    par(mfrow=c(1,2))
    plot(train$x, train$y, cex=0.5, pch=19, col="grey")
    title(main="k-nn fit: k=2")
    lines(train$x, true_model(train$x), col="black", lwd=2)
    lines(train$x, fit[,1], col="blue", lwd=2)
    plot(train$x, train$y, cex=0.5, pch=19, col="grey")
    title(main="k-nn fit: k=12")
    lines(train$x, true_model(train$x), col="black", lwd=2)
    lines(train$x, fit[,11], col="blue", lwd=2)
    return(k_vec)
  }
}


# Function to plot test set MSE
PlotTestError <- function(sim, base_plot_ls, base_plot_knn){
  linear=base_plot_ls
  k_vec=base_plot_knn
  train=sim$train
  test=sim$test
  testMSE_vec=numeric(length=length(k_vec))
  for (i in 1:length(k_vec)){
    near=kknn(y~x, train=train, test=test[,.(x)], k=k_vec[i], kernel='rectangular')
    testMSE_vec[i]=mean((near$fitted-test$y)^2)
  }
  lsMSE=mean((predict(linear, test)-test$y)^2)
  par(mfrow=c(1,1))
  plot(log(1/k_vec), testMSE_vec,
       ylim = c(0.95*min(c(testMSE_vec, lsMSE)),1.05*max(c(testMSE_vec, lsMSE))),
       type="n")
  title(main="Test set mean squared error")
  lines(log(1/k_vec), testMSE_vec, lty=6, lwd=3, col="forestgreen", type="o", cex=0.6 )
  abline(h=lsMSE, col="black", lty=2)
}
```

## Linear model (part 1.1 - 1.5)

Define function
```{r}
func_linear <- function(x){1.8*x+2}
```

Simulate data from the linear function

```{r}
sim_linear <- Simulate(func_linear, N_train, N_test)
```

Plot the training dataset, true relationship and best linear fit
```{r}
plot_ls<- Plot(sim_linear, method = "linear regression")
```

Plot k-NN fit with k=2 and k=12 respectively
```{r}
plot_knn <- Plot(sim_linear, method="k-nn")
```

Plot test set MSE

```{r}
plotMSE <- PlotTestError(sim_linear, plot_ls, plot_knn)
```

* Observation

Linear regression model outperforms k-NN algorithm for any value of $k$. In fact, when the true relationship is
linear, it is hard for a non-parametric approach to compete with linear regression. Also notice that when the
value of $k$ is large, i.e. the model is simple, k-NN performs only a little worse than least square regression. It
does far worse when $k$ is small.

## Almost linear model (part 1.6)

Define function
```{r}
func_almost_linear <- function(x){tanh(1.1*x)+2}
```

Simulate data and plot
```{r}
sim_aslinear <- Simulate(func_almost_linear, N_train, N_test)
plot_ls_aslinear <- Plot(sim_aslinear, method='linear regression')
plot_knn_aslinear <- Plot(sim_aslinear, method='k-nn')
plotMSE_aslinear <- PlotTestError(sim_aslinear, plot_ls_aslinear, plot_knn_aslinear)
```


* Observation

The test MSE for linear regression is superior to that of k-NN for low values of $k$. However, for larger $k$,
k-NN outperforms linear regression.



## Strongly nonlinear model (part 1.7)


Define function
```{r}
func_non_linear <- function(x){sin(2*x)+2}
```

Simulate data and plot
```{r}
sim_nlinear <- Simulate(func_non_linear, N_train, N_test)
plot_ls_nlinear <- Plot(sim_nlinear, method='linear regression')
plot_knn_nlinear <- Plot(sim_nlinear, method='k-nn')
plotMSE_nlinear <- PlotTestError(sim_nlinear, plot_ls_nlinear, plot_knn_nlinear)
```



* Observation

When the true relationship is highly non-linear, k-NN gives much better result than linear regression, for all
values of k. The test MSE of linear regression increases dramatically with the extent of non-linearity while
that of k-NN changes little.



## Generate data

```{r }
Xtrain=matrix(runif(N_train*20, -1, 1), nrow=N_train)
Xtest=matrix(runif(N_test*20, -1, 1), nrow=N_test)
Ytrain=sin(2*Xtrain[,1])+2+rnorm(N_train)
Ytest=sin(2*Xtest[,1])+2+rnorm(N_test)

for (pp in 1:5) {
  par(mfrow = c(1, 4), mai=c(0.2,0.2,0.2,0.2))
  for (p in seq((pp-1)*4+1, (pp)*4)) {
    
    Xtrain=matrix(runif(N_train*p, -1, 1), nrow=N_train)
    Xtest=matrix(runif(N_test*p, -1, 1), nrow=N_test)
    Ytrain=sin(2*Xtrain[,1])+2+0.3*rnorm(N_train)
    Ytest=sin(2*Xtest[,1])+2+0.3*rnorm(N_test)
    
    train=data.table(y=Ytrain, Xtrain[,1:p])
    colnames(train)=c("y", paste0("x", 1:p))
    test=data.table(y=Ytest, Xtest[,1:p])
    colnames(test)=c("y", paste0("x", 1:p))
    linear=lm(y~., data=train)
    
    lsMSE=mean((predict(linear, test)-test$y)^2)
    k_vec=seq(2, 15, 3)
    testMSE_vec=numeric(length=length(k_vec))
    for (i in 1:length(k_vec)){
      near=kknn(y~., train=train, test=test, k=k_vec[i], kernel='rectangular')
      testMSE_vec[i]=mean((near$fitted-test$y)^2)
    }
    plot(log(1/k_vec), testMSE_vec,
         ylim = c(0.,1.),
         type="n")
    title(main=paste("p = ",p))
    lines(log(1/k_vec), testMSE_vec, lty=6, lwd=1, col="forestgreen", type="o", cex=0.5 )
    abline(h=lsMSE, col="black", lty=2)
  }
}
```

* Observation

The increase in dimension only causes a small deterioration in the linear regression test set MSE, but it causes substantial increase in the MSE for k-NN. This is the curse-of-dimensionality problem, which results from the fact that in higher dimensions there is effectively a reduction in sample size. The general rule is linear regression will outperform k-NN method when there is a small number of observations per predictor, though in our simulation exercise this relationship does not hold monotonically.

## Optional bonus question

* Our conclusion concerning the relative performance of linear regression and k-NN as the extent of model non-linearity varies will ramain unchanged.

* We will have a wider range of values of k for which k-NN outperforms linear regression. In real world cases where the true relationship is barely linear, for any given p, an increase in the size of training sample boosts the number of sample points within each small neighborhood, which helps to improve the performance of k-NN. In other words, k-NN can now sustain more complex models without the peril of overfitting.

* Generally having a larger training set improves prediction accuracy in terms of test error by reducting the chance of overfitting.



# Question 2


```{r}
# set seed
set.seed(100)

# read data
download.file("https://github.com/ChicagoBoothML/MLClassData/raw/master/UsedCars/UsedCars.csv",
              "UsedCars.csv")
UsedCars <- read.csv(file="UsedCars.csv",head=TRUE,sep=",")

# splitting data into training and validation
N = dim(UsedCars)[1]
train_index = sample(N, size = N * 0.75, replace = FALSE)
train = UsedCars[train_index,]
test = UsedCars[-train_index, ]
```


## Part 3

```{r}
fit = lm(price ~ mileage, data = train)
plot(train$mileage, train$price, xlab = "mileage", ylab = "price", pch = 1, cex = 0.1)
abline(fit, col = "red", lwd = 2)
```

The linear regression can not capture the non-linear trend of the data.


## Part 4


```{r}
docv = function(x,y,set,nfold=10,doran=TRUE,verbose=TRUE,...)
{
  #a little error checking
  x = as.matrix(x)
  set = matrix(set, ncol = 1)
  if(!(is.matrix(x) | is.data.frame(x))) {cat('error in docv: x is not a matrix or data frame\n'); return(0)}
  if(!(is.vector(y))) {cat('error in docv: y is not a vector\n'); return(0)}
  if(!(length(y)==nrow(x))) {cat('error in docv: length(y) != nrow(x)\n'); return(0)}
  
  nset = nrow(set); 
  n=length(y) #get dimensions
  
  if(n==nfold) doran=FALSE #no need to shuffle if you are doing them all.
  cat('in docv: nset,n,nfold: ',nset,n,nfold,'\n')
  lossv = rep(0,nset) #return values
  if(doran) {ii = sample(1:n,n); y=y[ii]; x=x[ii,,drop=FALSE]} #shuffle rows
  
  fs = round(n/nfold) # fold size
  for(i in 1:nfold) { #fold loop
    bot = (i-1)*fs+1; 
    top = ifelse(i==nfold,n,i*fs); 
    ii = bot:top
    if(verbose) cat('on fold: ',i,', range: ',bot,':',top,'\n')
    xin = x[-ii,,drop=FALSE]; 
    yin=y[-ii]; 
    xout=x[ii,,drop=FALSE]; 
    yout=y[ii]
    xin = as.vector(xin)
    xout = as.vector(xout)
    datain = data.frame(x = xin, y = yin)
    dataout = data.frame(x = xout, y = yout)
    for(k in 1:nset) 
    { #setting loop
      fit = lm(y ~ poly(x, set[k,]), data = datain)
      yhat = predict(fit,  newdata = dataout)
      lossv[k]=lossv[k]+mean((yout-yhat)^2)
    } 
  } 
  return(lossv)
}



#### run cv
D = c(1,2,3,4,5,6, 7, 8, 9) # potential degrees
errors = docv(train$mileage, train$price, D, nfold = 10)
plot(D, errors, xlab = "highest degree", ylab = "Errors")
best.degree = D[which.min(errors)] # find the best degree

best.degree

fit = lm(price ~ poly(mileage, 7), data = train)
plot(train$mileage, train$price, xlab = "mileage", ylab = "price", pch = 1, cex = 0.1)
fitted = predict(fit, newdata = data.frame(mileage = sort(train$mileage)))
lines(sort(train$mileage),fitted,col="red",lwd=2, cex.lab=2)

# error
sqrt(mean((test$price -predict(fit, test))^2))
```


Due to randomness of the code, the result might be different for each
run.

## Part 5

```{r}
##########
## kNN
## Warning: super slow
##########

library(MASS)
library(kknn)
download.file("https://raw.githubusercontent.com/ChicagoBoothML/HelpR/master/docv.R", "docv.R")
source("docv.R") #this has docvknn used below
set.seed(99) #always set the seed! 

kv = 1:20 * 50   #these are the k values (k as in kNN) we will try
# since the possible k might be very large, we pick k every 50, from 50 to 1000

#does cross-validation for training data (x,y).

cv = docvknn(matrix(train$mileage,ncol=1),train$price,kv,nfold=10)
plot(kv, cv)
kbest = kv[which.min(cv)]

kfbest = kknn(price~mileage,train,data.frame(mileage=sort(train$mileage)),k=kbest,kernel = "rectangular")
plot(train$mileage, train$price, xlab = "mileage", ylab = "price", pch = 1, cex = 0.1)
lines(sort(train$mileage),kfbest$fitted,col="red",lwd=2, cex.lab=2)

# error
kfbest = kknn(price~mileage,train,test,k=kbest,kernel = "rectangular")
sqrt(mean((test$price - kfbest$fitted.values)^2))

##########
## Tree
##########

library(rpart)      # package for trees
library(rpart.plot) # package that enhances plotting capabilities for rpart
library(MASS)  # contains boston housing data


big.tree = rpart(price~mileage, data=train,
      control=rpart.control(minsplit=5,cp=0.0001,xval=10))

nbig = length(unique(big.tree$where))
cat('size of big tree: ',nbig,'\n')

(cptable = printcp(big.tree))
(bestcp = cptable[ which.min(cptable[,"xerror"]), "CP" ])   # this is the optimal cp parameter

plotcp(big.tree) # plot results

# show fit from some trees
oo = order(train$mileage)
cpvec = c(bestcp / 2, bestcp,.0157)

par(mfrow=c(3,2))
for(i in 1:3) {
  plot(train$mileage,train$price,pch=16,col='blue',cex=.5)
  ptree = prune(big.tree,cp=cpvec[i])
  pfit = predict(ptree)
  lines(train$mileage[oo],pfit[oo],col='red',lwd=2)
  title(paste('alpha = ',round(cpvec[i],3)))
  rpart.plot(ptree)
}

par(mfrow=c(1,1))
best.tree = prune(big.tree,cp=bestcp)
rpart.plot(best.tree)

plot(train$mileage, train$price, xlab = "mileage", ylab = "price", pch = 1, cex = 0.1)
lines(sort(train$mileage),predict(best.tree)[oo],col="red",lwd=2,cex.lab=2)

# error
sqrt(mean((test$price - predict(best.tree, test))^2))
```

From figures above, I would choose kNN because it’s robust and smooth. Tree and high-order polynomials
are less smooth and very sensitive to extreme values.


## Part 6

```{r}

###############################################################################
# 6
# Add one more variable

##########
## kNN
## Warning: super slow
##########



library(MASS)
library(kknn)
set.seed(99) #always set the seed! 

kv = 1:20 * 10  

cv = docvknn(scale(as.matrix(cbind(train$mileage, train$year))),train$price,kv,nfold=10)
plot(kv, cv)
kbest = kv[which.min(cv)]

train.scaled = train
train.scaled$year = scale(train.scaled$year)
train.scaled$mileage = scale(train.scaled$mileage)
test.scaled = test
test.scaled$year = scale(test.scaled$year)
test.scaled$mileage = scale(test.scaled$mileage)

kfbest = kknn(price~mileage+price,train.scaled,data.frame(mileage=sort(train.scaled$mileage)),k=kbest,kernel = "rectangular")
plot(train.scaled$mileage, train$price, xlab = "mileage", ylab = "price", pch = 1, cex = 0.1)
lines(sort(train.scaled$mileage),kfbest$fitted,col="red",lwd=2, cex.lab=2)



# error
kfbest = kknn(price~mileage,train.scaled,test.scaled,k=kbest,kernel = "rectangular")
sqrt(mean((test.scaled$price - kfbest$fitted.values)^2))



##########
## Tree
##########
library(rpart)      # package for trees
library(rpart.plot) # package that enhances plotting capabilities for rpart
library(MASS)  # contains boston housing data


big.tree = rpart(price~mileage+price, data=train,
                 control=rpart.control(minsplit=5,cp=0.0001,xval=10))

nbig = length(unique(big.tree$where))
cat('size of big tree: ',nbig,'\n')

(cptable = printcp(big.tree))
(bestcp = cptable[ which.min(cptable[,"xerror"]), "CP" ])   # this is the optimal cp parameter

plotcp(big.tree) # plot results

# show fit from some trees
oo = order(train$mileage)
cpvec = c(bestcp / 2, bestcp,.0157)

par(mfrow=c(3,2))
for(i in 1:3) {
  plot(train$mileage,train$price,pch=16,col='blue',cex=.5)
  ptree = prune(big.tree,cp=cpvec[i])
  pfit = predict(ptree)
  lines(train$mileage[oo],pfit[oo],col='red',lwd=2)
  title(paste('alpha = ',round(cpvec[i],3)))
  rpart.plot(ptree)
}

par(mfrow=c(1,1))
best.tree = prune(big.tree,cp=bestcp)
rpart.plot(best.tree)

plot(train$mileage, train$price, xlab = "mileage", ylab = "price", pch = 1, cex = 0.1)
lines(sort(train$mileage),predict(best.tree)[oo],col="red",lwd=2,cex.lab=2)

# error
sqrt(mean((test$price - predict(best.tree, test))^2))

```

Comparing with one variable case, now k of kNN and optimal tree size are smaller. Both method get lower RMSE with one more variable.

## Part 7


```{r}
big.tree = rpart(price~., data=train,
                 control=rpart.control(minsplit=5,cp=0.0001,xval=10))

nbig = length(unique(big.tree$where))
cat('size of big tree: ',nbig,'\n')

(cptable = printcp(big.tree))
(bestcp = cptable[ which.min(cptable[,"xerror"]), "CP" ])   # this is the optimal cp parameter

plotcp(big.tree) # plot results

# show fit from some trees
oo = order(train$mileage)
cpvec = c(bestcp / 2, bestcp,.0157)

par(mfrow=c(3,2))
for(i in 1:3) {
  plot(train$mileage,train$price,pch=16,col='blue',cex=.5)
  ptree = prune(big.tree,cp=cpvec[i])
  pfit = predict(ptree)
  lines(train$mileage[oo],pfit[oo],col='red',lwd=2)
  title(paste('alpha = ',round(cpvec[i],3)))
  rpart.plot(ptree)
}

par(mfrow=c(1,1))
best.tree = prune(big.tree,cp=bestcp)
rpart.plot(best.tree)

plot(train$mileage, train$price, xlab = "mileage", ylab = "price", pch = 1, cex = 0.1)
lines(sort(train$mileage),predict(best.tree)[oo],col="red",lwd=2,cex.lab=2)

# error
sqrt(mean((test$price - predict(best.tree, test))^2))
```

## Bonus


One possible procedure

* For each fixed $k$, there are $p \choose k$ possible k-variable models. Run cross validation and compute RMSE for all possible models. The model with smallest RMSE is the optimal k-variable model.

* For all $k = 1, \ldots, p$. Compare RMSE of all optimal k-variable models and find the one with smallest RMSE.

